---
title: "SCP exam"
author: "Oussama"
date: "2025-01-24"
output: pdf_document
---

```{r}
library(tidyverse)

df <- read_csv("cancer issue.csv")
df

```

# Part 1: Exploratory Data Analysis

## Question 1: What is the distribution of cancer types?

```{r}
distribution_cancer_type <- df %>% 
  count(CancerType) %>%
  arrange(desc(n))

print(distribution_cancer_type)

ggplot(distribution_cancer_type, aes(x = reorder(CancerType,-n),y=n )) +
  geom_col(width = 0.5)+
  geom_label(aes(label = n),hjust = 1)+
  coord_flip()+
  xlab("Cancer type") +
  ylab("count") + 
  labs(title = "distribution of cancer types")
   

```

## Question 2: What is the average BMI by cancer stage?

```{r}
 df %>% 
  group_by(Stage) %>% 
  summarize(mean_bmi = mean(BMI))
```

## Question 3: What is the survival distribution by hospital region?

```{r}
df %>%
  group_by(HospitalRegion) %>% 
  count(SurvivalMonths) %>% 
  summarise(count = sum(n))
  
```

## Question 4: What is the tumor size distribution?

```{r}
df %>% 
  group_by(TumorSize) %>% 
  count(TumorSize) %>%
  arrange(desc(TumorSize))
```

## Question 5: What is the relationship between tumor size and survival months?

```{r}
df %>% select(TumorSize,SurvivalMonths) %>% 
  summarise(correlation = cor(SurvivalMonths,TumorSize))

if (correlation > 0) {
  paste("There is a positive relationship between Tumor Size and Survival Months the correlation coefficient is ",correlation)
} else if (correlation < 0) {
  paste("There is a negative relationship between Tumor Size and Survival Months correlation coefficient is ",correlation)
} else {
  print("There is no relationship between TumorSize and SurvivalMonths")
}
```

## Question 6: What is the proportion of smokers in the dataset?

```{r}
smokers_prop <- df %>% filter(SmokingStatus == "Smoker") %>% 
  summarise(p = (n()/nrow(df))*100)

paste("the proportion or smokers in the dataset is ",smokers_prop,"%")
```

## Question 7: What is the average tumor size by cancer type?

```{r}
df %>% group_by(CancerType) %>% 
  summarise(avg_tumorsize = mean(TumorSize))
```

## Question 8: Is there a difference in average BMI between smokers and non-smokers?


```{r}
smok <- df %>% filter(SmokingStatus == "Smoker") %>% 
  summarise(avgbmi = mean(BMI))

non_smok <- df %>% filter(SmokingStatus == "Non-Smoker") %>% 
  summarise(avgbm = mean(BMI))

print(smok==non_smok)
```

## Question 9: What is the distribution of cancer stages?

```{r}
distrib_cancer_stage <- df %>% 
  count(Stage) %>%
  arrange(desc(n))


ggplot(distrib_cancer_stage, aes(x = reorder(Stage,-n),y=n )) +
  geom_col(width = 0.5)+
  geom_label(aes(label = n),hjust = 1)+
  xlab("Cancer type") +
  ylab("count") + 
  labs(title = "distribution of cancer types")
```

## Question 10: How many patients are in each hospital region?

```{r}
df %>% group_by(HospitalRegion) %>% 
  count()
```

## Question 11: What is the survival rate by cancer type?

```{r}
df %>% 
  group_by(CancerType) %>% 
  summarise(Proportion = (sum(Recurrence == "Yes") / n())*100)
```

## Question 12: What is the relationship between age and survival months?

```{r}
df %>% select(Age,SurvivalMonths) %>% 
  summarise(correlation = cor(Age,SurvivalMonths))

if (correlation > 0) {
  paste("There is a positive relationship between age and Survival Months the correlation coefficient is ",correlation)
} else if (correlation < 0) {
  paste("There is a negative relationship between age and Survival Months correlation coefficient is ",correlation)
} else {
  print("There is no relationship between and SurvivalMonths")
}
```

## Question 13: What is the most common treatment by cancer type?

```{r}
df %>% group_by(CancerType) %>% 
  count(TreatmentType) 
```

## Question 14: What is the gender distribution of patients?

```{r}
df %>% group_by(Gender) %>% 
  count() 

            
```

## Question 15: How does smoking status vary by hospital region?

```{r}
df %>% group_by(HospitalRegion,SmokingStatus) %>% 
  count(SmokingStatus)
  
```

## Question 16: What is the average survival months by genetic mutation type?

```{r}
df %>% group_by(GeneticMarker) %>% 
  summarise(avg = mean(SurvivalMonths))
```

## Question 17: What is the distribution of patients by age group?

```{r}
df %>% group_by(Age) %>% 
  count()
```

## Question 18: What are the most common combinations of cancer type and stage?

```{r}
df %>% group_by(CancerType,Stage) %>% 
  count() %>% 
  arrange(desc(n)) %>% 
  head(5)
```

## Question 19: What is the relationship between BMI and tumor size?

```{r}
df %>% select(BMI,TumorSize) %>% 
  summarise(correlation = cor(BMI,TumorSize))

if (correlation > 0) {
  paste("There is a positive relationship between bmi and tumor size the correlation coefficient is ",correlation)
} else if (correlation < 0) {
  paste("There is a negative relationship between bmi and tumor size correlation coefficient is ",correlation)
} else {
  print("There is no relationship between bmi and tumor size")
}
```

## Question 20: How many patients have a family history of cancer?

```{r}
hist <- df %>% filter(FamilyHistory == "Yes") %>% 
  count()

paste("the number of patients that have a family history of cancer is ", hist)
```

#Part 2: Function Creation and Error Handling
## Question 1: Create a function to classify age groups 

```{r}
age_groups <- function(age) {
  if (age < 18) {
    return("Child/teen")
  } else if (age >= 18 & age < 30) {
    return("Young adult")
  } else if(age >= 30 & age <60 ){
    return("Adult")
  }
  else{
    return("Senior")
  }
}

df <- df %>% mutate(AgeGroup = sapply(Age, age_groups))

```

## Question 2: Create a function to normalize tumor size.

```{r}
normalize <- function(tumor_size) {
  mean_size <- mean(tumor_size)
  ecarttype <- sd(tumor_size)
  return((tumor_size - mean_size) / ecarttype)
}

df <- df %>% mutate(norm_tumorsize = normalize(TumorSize))

df$norm_tumorsize


```

## Question 17Part 1: What is the distribution of patients by age group?

```{r}
df %>% group_by(AgeGroup) %>% 
  count() %>% 
  arrange(desc(n))
```

## Question 3: Create a function to calculate BMI categories

```{r}
bmi_category <- function(bmi) {
  if (bmi < 18.5) {
    return("Underweight")
  } else if (bmi >= 18.5 & bmi < 25) {
    return("Normal")
  } else if (bmi >= 25 & bmi < 30) {
    return("Overweight")
  } else {
    return("Obese")
  }
}

df <- df %>% mutate(BMICategory = sapply(BMI,bmi_category))

```

## Question 4: Write a function to check if survival months are valid (non-negative).

```{r}
check_survmonths <- function(survival_months) {
  if (survival_months < 0) {
    return("Invalid")
  } else {
    return("Valid")
  }
}

df <- df %>% mutate(SurvivalMonthsValidity = sapply(SurvivalMonths, check_survmonths))

```

## Question 5: Develop a function to calculate tumor size as a percentage of maximum size.

```{r}
tumorsize_percentage <- function(tumor_size) {
  max_size <- max(tumor_size)
  return((tumor_size /max_size) * 100)
}

df <- df %>% mutate(TumorSizePercentage = tumorsize_percentage(TumorSize))

```

## Question 6: Write a function to generate descriptive statistics for a numeric column

```{r}
descriptive_stats <- function(var) {
  statistics <-list(
    Mean = mean(var),
    Median = median(var),
    Min = min(var),
    Max = max(var))
  return(statistics)
}
descriptive_stats(df$TumorSize)
```

## Question 7: Create a function to filter data by cancer type and stage.

```{r}
type_stage <- function(data, cancer_type, stage) {
  filtered_data <- data %>% filter(CancerType == cancer_type & Stage == stage)
  return(filtered_data)
}

type_stage(df, "Breast", "II")
```

## Question 8: Write a function to compute the average survival months for each region.

```{r}
survival_region <- function() {
  df %>% group_by(HospitalRegion) %>% 
  summarise(AvgSurvivalMonths = mean(SurvivalMonths))
}

survival_region()
```



## Question 9: Develop a function to create a summary table for a given variable 

```{r}
summary_table <- function(variable) {
  df %>% count(!!sym(variable)) %>%
    arrange(desc(n))
}
variablee = "CancerType"
summary_table(variablee)

```

## Question 10: Write a function to identify patients with above-average tumor sizes

```{r}
above_avgtumorsize <- function() {
  avg_size <- mean(df$TumorSize)
  df %>%
  filter(TumorSize > avg_size)
}

above_avgtumorsize()
```

## Question 11: Create a function to categorize survival months into short, medium, and long terms


```{r}
categorize_survival <- function(survival_months) {
  if (survival_months < 12) {
    return("Short Term")
  } else if (survival_months >= 12 & survival_months < 60) {
    return("Medium Term")
  } else {
    return("Long Term")
  }
}

df <- df %>% mutate(SurvivalCategory = sapply(SurvivalMonths, categorize_survival))

```
## Question 12: Develop a function to check for missing values in each column

```{r}
check_missing_values <- function() {
  colSums(is.na(df))
}

check_missing_values()
```

## Question 13: Write a function to calculate the median tumor size for smokers and non-smokers.

```{r}
median_tumorsize <- function() {
  df %>% group_by(SmokingStatus) %>%     summarise(MedianTumorsize = median(TumorSize))
}

median_tumorsize()
```

## Question 14: Create a function to count the number of patients by gender and cancer type.

```{r}
count_gender <- function() {
  df %>% group_by(Gender, CancerType) %>% summarise(Count = n())
}

count_gender()
```

## Question 15: Write a function to group data by region and calculate average BMI.

```{r}
avg_bmi_region <- function() {
  df %>% group_by(HospitalRegion) %>% summarise(AvgBMI = mean(BMI))
}

avg_bmi_region()
```

## Question 16: Develop a function to identify patients with genetic mutations linked to survival rates.

```{r}
genetic_survival <- function() {
  df %>% filter(GeneticMarker != "None") %>% group_by(GeneticMarker) %>% summarise(AvgSurvivMonths = mean(SurvivalMonths))
}

genetic_survival()
```

## Question 17: Write a function to summarize treatment response rates by cancer stage.

```{r}
treatment_response <- function() {
  df %>% group_by(Stage, TreatmentResponse) %>% summarise(Count = n())
}

treatment_response()
```

## Question 18:	Create a function to check for outliers in numeric columns.

```{r}
check_outliers <- function(column) {
  mean <- mean(column)
  median <- median(column)
  
  down <- mean - median
  up <- mean + median
  
  df %>% filter(column < down | column > up)
}
numeric_col <- df$TumorSize
check_outliers(numeric_col)
# no outliers in this case
```

## Question 19: Write a function to calculate the proportion of patients with recurrence by region.

```{r}
recurrence_region <- function() {
  df %>% group_by(HospitalRegion) %>% summarise(RecurrenceRate = mean(Recurrence == "Yes"))
}

recurrence_region()
```

## Question 20: Develop a function to flag invalid BMI values (e.g., below 10 or above 60).

```{r}
flag_bmi <- function(bmi) {
  if (bmi < 10 | bmi > 60) {
    return("Invalid")
  } else {
    return("Valid")
  }
}

df <- df %>% mutate(BMIValidity = sapply(BMI, flag_bmi))
```




